---
title: "ERCOT Ancillary Services Price Forecast"
author: "Skyler Ju, Joy Wu, Judy Hua Zhu"
output: pdf_document
#always_allow_html: true
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: sentence
---

## Setting R code chunk options

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

```{r package, message=FALSE, warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(zoo)
library(readxl)
```

## Importing data

The dataset contains ERCOT ancillary services Market Clearing Prices for Capacity (MCPC) from December 2023 through April 2026. The raw data is in long format with one row per delivery date, hour ending, and ancillary service type combination. The five ancillary service types are ECRS, NSPIN, REGDN, REGUP, and RRS.

```{r}
#Importing raw data
raw_data <- read.csv("./Data/Raw_data.csv")

#Inspect data
head(raw_data, 10)
str(raw_data)
nobs <- nrow(raw_data)
summary(raw_data)
```
## Wrangling and Processing the Dataset

The goal is to reshape the data so that each row represents a unique delivery date and hour ending, with the five ancillary service types as separate columns containing their MCPC values. We also need to handle DST (Day Light Saving) flag entries where fall-back days have a duplicate 02:00 hour.

### Step 1: Handle DST duplicate hours

Fall-back days (November) have two 02:00 entries per ancillary type, one with DSTFlag = true and one with DSTFlag = false. We average the two to collapse them into a single 02:00 entry per day.

```{r wrangle_dst, message=FALSE, warning=FALSE}
#Check DST flagged rows
cat("Number of DST flagged rows:", sum(raw_data$DSTFlag == "true"), "\n")
raw_data %>% filter(DSTFlag == "true") %>% head(10)


#Average MCPC for duplicate DST hours, then drop DSTFlag column
data_clean <- raw_data %>%
  mutate(deliveryDate = ymd(deliveryDate)) %>%
  rename(Date = deliveryDate) %>%
  group_by(Date, hourEnding, ancillaryType) %>%
  summarise(MCPC = mean(MCPC, na.rm = TRUE), .groups = "drop")

cat("Rows after DST handling:", nrow(data_clean), "\n")
cat("Expected (848 days x 24 hours x 5 types):", 
    n_distinct(data_clean$Date) * 24 * 5, "\n")
```
> * We want to flag here that the result prints: "Rows after DST handling: 101745 
Expected (848 days x 24 hours x 5 types): 101760". The difference is 15, which is 3 missing hours times 5 ancillary types. Those are the three spring-forward days (March 2024, March 2025, March 2026) where the 02:00 hour doesn't exist because the clock jumps from 1:00 AM to 3:00 AM. The above function would already handle this situation, but it can't create rows that don't exist in the raw data. Those missing 02:00 entries will show up as NAs after the pivot step, and the na.approx() function will fill them in by interpolating between the 01:00 and 03:00 values. 

### Step 2: Pivot ancillary service types to columns

```{r wrangle_pivot, message=FALSE, warning=FALSE}
#Pivot wider so each ancillary type becomes its own MCPC column
data_wide <- data_clean %>%
  pivot_wider(names_from = ancillaryType, 
              values_from = MCPC)

#Inspect the pivoted data
head(data_wide)
str(data_wide)

#Check for NAs after pivot (missing hours from spring-forward DST)
cat("\nNAs per column:\n")
colSums(is.na(data_wide))
```

### Step 3: Handle missing hours from spring-forward DST

The pivot created rows only for date-hour combinations that exist in the raw data. Spring-forward days (March) are missing the 02:00 hour entirely because clocks jump from 1:00 AM to 3:00 AM, so those rows are absent rather than NA. We need to add them back as explicit rows so the time series has evenly spaced hourly observations, then fill the gaps using interpolation.

```{r wrangle_spring_dst, message=FALSE, warning=FALSE}
#Create a complete grid of all dates and hours
full_data <- expand.grid(
  Date = seq(min(data_wide$Date), max(data_wide$Date), by = "day"),
  hourEnding = unique(data_wide$hourEnding)
) %>%
  arrange(Date, hourEnding)

#Join to fill in missing spring-forward hours as NAs
data_wide <- full_data %>%
  left_join(data_wide, by = c("Date", "hourEnding"))

#Check NAs now
cat("NAs per column:\n")
colSums(is.na(data_wide))

#Fill spring-forward NAs using interpolation within each date
data_wide <- data_wide %>%
  group_by(Date) %>%
  mutate(across(c(ECRS, NSPIN, REGDN, REGUP, RRS), 
                ~na.approx(., na.rm = FALSE))) %>%
  ungroup()

#Verify
cat("Rows after filling:", nrow(data_wide), "\n")
cat("Expected (848 x 24):", n_distinct(data_wide$Date) * 24, "\n")
cat("NAs remaining:", sum(is.na(data_wide %>% select(ECRS:RRS))), "\n")
```


### Step 4: Sort by date and hour

```{r wrangle_sort, message=FALSE, warning=FALSE}
#Sort chronologically
data_wide <- data_wide %>%
  arrange(Date, hourEnding)

head(data_wide, 25)
tail(data_wide, 25)
summary(data_wide)
```



## Initial Plots

```{r initial_plots, message=FALSE, warning=FALSE}
#For plotting, create a datetime column combining date and hour
data_plot <- data_wide %>%
  mutate(hour_num = as.numeric(substr(hourEnding, 1, 2)),
         DateTime = Date + hours(hour_num))

#Plot all five ancillary types over time
#Using daily averages for cleaner visualization
daily_avg <- data_plot %>%
  group_by(Date) %>%
  summarise(across(c(ECRS, NSPIN, REGDN, REGUP, RRS), 
                   ~mean(., na.rm = TRUE))) %>%
  pivot_longer(cols = c(ECRS, NSPIN, REGDN, REGUP, RRS),
               names_to = "Type",
               values_to = "MCPC")

ggplot(daily_avg, aes(x = Date, y = MCPC, color = Type)) +
  geom_line(alpha = 0.7) +
  ylab("Daily Average MCPC ($/MW)") +
  xlab("Date") +
  ggtitle("ERCOT Ancillary Services MCPC: All Types") +
  theme(legend.position = "bottom")

#Individual plots for each ancillary type
for(type in c("ECRS", "NSPIN", "REGDN", "REGUP", "RRS")) {
  p <- daily_avg %>%
    filter(Type == type) %>%
    ggplot(aes(x = Date, y = MCPC)) +
    geom_line(color = "blue") +
    ylab("Daily Average MCPC ($/MW)") +
    xlab("Date") +
    ggtitle(paste("ERCOT", type, "MCPC"))
  print(p)
}
```

## Creating Time Series Objects

Since this is hourly data with multiple seasonal components (daily cycle of 24 hours and weekly cycle of 168 hours), we use `msts()`. We create a time series object for each ancillary service type.

```{r ts_objects, message=FALSE, warning=FALSE}
#Create msts objects for each ancillary type
#Seasonal periods: 24 (daily), 168 (weekly = 24*7)
ts_ECRS <- msts(data_wide$ECRS,
                seasonal.periods = c(24, 168),
                start = c(2023, 347, 1))
                #day 347 = Dec 13

ts_NSPIN <- msts(data_wide$NSPIN,
                 seasonal.periods = c(24, 168),
                 start = c(2023, 347, 1))

ts_REGDN <- msts(data_wide$REGDN,
                 seasonal.periods = c(24, 168),
                 start = c(2023, 347, 1))

ts_REGUP <- msts(data_wide$REGUP,
                 seasonal.periods = c(24, 168),
                 start = c(2023, 347, 1))

ts_RRS <- msts(data_wide$RRS,
               seasonal.periods = c(24, 168),
               start = c(2023, 347, 1))

cat("Length of each ts object:", length(ts_ECRS), "hourly observations\n")
cat("Expected:", nrow(data_wide), "\n")
```

## Decomposing the Time Series

Using `mstl()` to decompose the multi-seasonal time series and inspect the daily and weekly seasonal components. Starting with ECRS as an example.

```{r decompose, message=FALSE, warning=FALSE}
#Decompose ECRS
ts_ECRS %>% mstl() %>%
  autoplot() +
  ggtitle("Decomposition of ECRS MCPC")

#Decompose RRS
ts_RRS %>% mstl() %>%
  autoplot() +
  ggtitle("Decomposition of RRS MCPC")

#Decompose REGUP
ts_REGUP %>% mstl() %>%
  autoplot() +
  ggtitle("Decomposition of REGUP MCPC")

#Decompose REGDN
ts_REGDN %>% mstl() %>%
  autoplot() +
  ggtitle("Decomposition of REGDN MCPC")

#Decompose RRS
ts_RRS %>% mstl() %>%
  autoplot() +
  ggtitle("Decomposition of RRS MCPC")
```

## ACF and PACF

```{r acf_pacf, message=FALSE, warning=FALSE}
#ACF and PACF for ECRS
par(mfrow=c(1,2))
Acf(data_wide$ECRS, lag.max = 168, 
    main = "ACF - ECRS MCPC")
Pacf(data_wide$ECRS, lag.max = 168, 
     main = "PACF - ECRS MCPC")
par(mfrow=c(1,1))

#ACF and PACF for RRS
par(mfrow=c(1,2))
Acf(data_wide$RRS, lag.max = 168, 
    main = "ACF - RRS MCPC")
Pacf(data_wide$RRS, lag.max = 168, 
     main = "PACF - RRS MCPC")
par(mfrow=c(1,1))

#ACF and PACF for REGUP
par(mfrow=c(1,2))
Acf(data_wide$REGUP, lag.max = 168, 
    main = "ACF - REGUP MCPC")
Pacf(data_wide$REGUP, lag.max = 168, 
     main = "PACF - REGUP MCPC")
par(mfrow=c(1,1))

#ACF and PACF for REGDN
par(mfrow=c(1,2))
Acf(data_wide$REGDN, lag.max = 168, 
    main = "ACF - REGDN MCPC")
Pacf(data_wide$REGUP, lag.max = 168, 
     main = "PACF - REGDN MCPC")
par(mfrow=c(1,1))

#ACF and PACF for RRS
par(mfrow=c(1,2))
Acf(data_wide$RRS, lag.max = 168, 
    main = "ACF - RRS MCPC")
Pacf(data_wide$REGUP, lag.max = 168, 
     main = "PACF - RRS MCPC")
par(mfrow=c(1,1))
```

## Setting Up Training and Test Sets

Training period: start of dataset (December 13, 2023) through December 31, 2025. Test period: January 1, 2026 through the end of the dataset (April 8, 2026).

```{r train_test_split, message=FALSE, warning=FALSE}
#Count hours in training period (Dec 2023 - Dec 2025)
n_train <- sum(data_wide$Date <= as.Date("2025-12-31"))

#Count hours in test period (Jan 2026 - Apr 2026)
n_test <- sum(data_wide$Date >= as.Date("2026-01-01"))

cat("Training observations (hours):", n_train, "\n")
cat("Test observations (hours):", n_test, "\n")
cat("Training days:", n_train / 24, "\n")
cat("Test days:", n_test / 24, "\n")

#Create training and test subsets for each ancillary type
#ECRS
ts_ECRS_train <- subset(ts_ECRS, end = n_train)
ts_ECRS_test  <- subset(ts_ECRS, start = n_train + 1)

#NSPIN
ts_NSPIN_train <- subset(ts_NSPIN, end = n_train)
ts_NSPIN_test  <- subset(ts_NSPIN, start = n_train + 1)

#REGDN
ts_REGDN_train <- subset(ts_REGDN, end = n_train)
ts_REGDN_test  <- subset(ts_REGDN, start = n_train + 1)

#REGUP
ts_REGUP_train <- subset(ts_REGUP, end = n_train)
ts_REGUP_test  <- subset(ts_REGUP, start = n_train + 1)

#RRS
ts_RRS_train <- subset(ts_RRS, end = n_train)
ts_RRS_test  <- subset(ts_RRS, start = n_train + 1)

#Verify lengths
cat("\nVerification:\n")
cat("ECRS train:", length(ts_ECRS_train), 
    "test:", length(ts_ECRS_test), 
    "total:", length(ts_ECRS_train) + length(ts_ECRS_test), "\n")

#Plot train and test (horizontal axis has cumulative units)
autoplot(ts_ECRS_train) +
  ylab("MCPC ($/MW)") +
  ggtitle("ECRS Training Data: Dec 2023 - Dec 2025")

autoplot(ts_ECRS_test) +
  ylab("MCPC ($/MW)") +
  ggtitle("ECRS Test Data: Jan 2026 - Apr 2026")

autoplot(ts_NSPIN_train) +
  ylab("MCPC ($/MW)") +
  ggtitle("NSPIN Training Data: Dec 2023 - Dec 2025")

autoplot(ts_NSPIN_test) +
  ylab("MCPC ($/MW)") +
  ggtitle("NSPIN Test Data: Jan 2026 - Apr 2026")

autoplot(ts_REGDN_train) +
  ylab("MCPC ($/MW)") +
  ggtitle("REGDN Training Data: Dec 2023 - Dec 2025")

autoplot(ts_REGDN_test) +
  ylab("MCPC ($/MW)") +
  ggtitle("REGDN Test Data: Jan 2026 - Apr 2026")

autoplot(ts_REGUP_train) +
  ylab("MCPC ($/MW)") +
  ggtitle("REGUP Training Data: Dec 2023 - Dec 2025")

autoplot(ts_REGUP_test) +
  ylab("MCPC ($/MW)") +
  ggtitle("REGUP Test Data: Jan 2026 - Apr 2026")

autoplot(ts_RRS_train) +
  ylab("MCPC ($/MW)") +
  ggtitle("RRS Training Data: Dec 2023 - Dec 2025")

autoplot(ts_RRS_test) +
  ylab("MCPC ($/MW)") +
  ggtitle("RRS Test Data: Jan 2026 - Apr 2026")




#Plot train and test using data frame with actual dates
data_plot <- data_wide %>%
  mutate(hour_num = as.numeric(substr(hourEnding, 1, 2)),
         DateTime = Date + hours(hour_num),
         Period = ifelse(Date <= as.Date("2025-12-31"), 
                         "Train", "Test"))

ggplot(data_plot, aes(x = DateTime, y = ECRS, color = Period)) +
  geom_line(alpha = 0.5) +
  ylab("MCPC ($/MW)") +
  xlab("Date") +
  ggtitle("ECRS MCPC: Training vs Test Period") +
  scale_x_datetime(date_breaks = "3 months", date_labels = "%b %Y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

for(type in c("ECRS", "NSPIN", "REGDN", "REGUP", "RRS")) {
  p <- ggplot(data_plot, aes(x = DateTime, y = .data[[type]], color = Period)) +
    geom_line(alpha = 0.5) +
    ylab("MCPC ($/MW)") +
    xlab("Date") +
    ggtitle(paste(type, "MCPC: Training vs Test Period")) +
    scale_x_datetime(date_breaks = "3 months", date_labels = "%b %Y") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p)
}

```

