---
output:
  pdf_document: default
  html_document: default
---
## Setting R code chunk options

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages and initializing

```{r package, message=FALSE, warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)
library(tidyverse)
```

## Loading the Processed Training Data

```{r load_data, message=FALSE, warning=FALSE}
#Load processed training data exported from data exploration
train_data <- read.csv("../Data/train_period_clean/data_train.csv")

cat("Training rows:", nrow(train_data),
    "(", n_distinct(train_data$Date), "days )\n")
```

## Creating Training Period Time Series Object

```{r ts_object}
ts_RRS_arima <- ts(train_data$RRS,
                   frequency = 24,
                   start = c(1, 1))

cat("Training ts length:", length(ts_RRS_arima), "hourly observations\n")
```

## ARIMA Models

We include three ARIMA-based models: a baseline ARIMA(1,1,1), an ARIMA(1,1,1) with Fourier terms at k=6, and an ARIMA(1,1,1) with Fourier terms at the optimal k selected through grid search. Model order was chosen by inspecting the differenced ACF and PACF series.

#### Model 1: ARIMA(1,1,1)

**Differenced Series: ACF and PACF**

The data exploration section showed that the RRS series is non-stationary, so we applied one round of first differencing before inspecting the ACF and PACF to identify the model order.

```{r acf_pacf_diff, message=FALSE, warning=FALSE, fig.width=14, fig.height=6}
#Apply first difference
RRS_diff1 <- diff(train_data$RRS, differences = 1)

par(mfrow=c(1,2))
Acf(RRS_diff1,  lag.max = 168, main = "ACF - First-Differenced RRS")
Pacf(RRS_diff1, lag.max = 168, main = "PACF - First-Differenced RRS")
par(mfrow=c(1,1))
```
After differencing, the slow ACF decay in the original series is gone, confirming d=1 is sufficient for stationarity. The ACF shows a significant negative spike at lag 1 followed by values mostly within the confidence bounds, consistent with MA(1), so we set q=1. The PACF shows a dominant spike at lag 1 followed by smaller significant spikes at higher lags, so we set p=1. Spikes at lags 24 and 48 are also visible in the ACF, pointing to a residual daily seasonal cycle that first differencing does not remove. It is addressed through Fourier terms in the next step.

**Model Fit and Residual Diagnostics**

```{r fit_arima}
ARIMA_RRS <- Arima(ts_RRS_arima,
                   order = c(1, 1, 1),
                   include.drift = FALSE)

print(ARIMA_RRS)
```

```{r residuals, message=FALSE, warning=FALSE, fig.width=14, fig.height=8}
checkresiduals(ARIMA_RRS)
```

```{r ljung_box}
lb_test <- Box.test(ARIMA_RRS$residuals, lag = 24, type = "Ljung-Box")
print(lb_test)
```
The Ljung-Box test strongly rejects white noise (Q*=3546.2, p<2.2e-16), showing significant autocorrelation remains in the residuals. The residual ACF shows clear spikes at lags 24 and 48, confirming that the daily seasonal cycle is not captured by ARIMA(1,1,1) alone. The histogram is heavily right skewed, reflecting the non-normal distribution driven by extreme price events. These results motivate adding Fourier terms to directly model the daily seasonal cycle in the next step.

#### Model 2: ARIMA(1,1,1) + Fourier (k = 6)

The residual diagnostics from ARIMA(1,1,1) show that the daily seasonal cycle at lag 24 is not captured by the base model. We extend the model by adding Fourier terms as external regressors. For a daily cycle of period 24, k ranges from 1 to 12 (period/2). We set k = 6 as a balanced choice between underfitting and overfitting.

**Model Fit and Residual Diagnostics**

```{r fourier_fit, message=FALSE, warning=FALSE}
#Generate Fourier terms for daily cycle (period = 24, k = 6)
fourier_train <- fourier(ts_RRS_arima, K = 6)

#Fit ARIMA(1,1,1) with Fourier terms as external regressors
ARIMA_Fourier_RRS <- Arima(ts_RRS_arima,
                            order = c(1, 1, 1),
                            include.drift = FALSE,
                            xreg = fourier_train)

print(ARIMA_Fourier_RRS)
```

```{r fourier_residuals, message=FALSE, warning=FALSE, fig.width=14, fig.height=8}
checkresiduals(ARIMA_Fourier_RRS)
```

```{r fourier_ljung_box}
lb_test_fourier <- Box.test(ARIMA_Fourier_RRS$residuals, lag = 24, type = "Ljung-Box")
print(lb_test_fourier)
```
The AIC drops from 159532.4 to 159218.3, showing better overall fit. However, the Ljung Box test still strongly rejects white noise (Q*=2810.1, p<2.2e-16), and spikes at lags 24 and 48 remain in the residual ACF, suggesting the daily seasonal pattern in RRS prices is too irregular to be fully captured by fixed Fourier.

**Accuracy on Test Period**
```{r}
#Load test data
test_data <- read.csv("../Data/test_period_clean/data_test.csv")
n_test <- nrow(test_data)

#ARIMA(1,1,1) forecast and accuracy
forecast_arima <- forecast(ARIMA_RRS, h = n_test)
accuracy(forecast_arima, test_data$RRS)

```

```{r}
#ARIMA + Fourier forecast and accuracy
fourier_test <- fourier(ts_RRS_arima, K=6, h=n_test)
forecast_fourier <- forecast(ARIMA_Fourier_RRS, xreg=fourier_test, h=n_test)
accuracy(forecast_fourier, test_data$RRS)
```
The Fourier extension marginally improves training set RMSE (20.15 vs. 20.34 for the baseline), but this does not carry over to the test set. Test RMSE increases slightly from 29.94 to 30.06, and test MAPE jumps significantly from 180.6 to 504.1, suggesting K = 6 overfits the training data and does not generalize well to the test period. This motivates the grid search for a better-fitting k.

#### Model 3: ARIMA(1,1,1) + Fourier (k=2)

**Fitting ARIMA + Fourier (grid search most suiatble K)**
```{r}
K_values <- 1:12

results <- lapply(K_values, function(K) {
  
  fourier_train <- fourier(ts_RRS_arima, K = K)
  
  model <- Arima(ts_RRS_arima,
                 order = c(1,1,1),
                 xreg = fourier_train)
  
  n_test <- nrow(test_data)
  fourier_test <- fourier(ts_RRS_arima, K = K, h = n_test)
  
  fc <- forecast(model, xreg = fourier_test, h = n_test)
  
  acc <- accuracy(fc, test_data$RRS)
  
  data.frame(
    K = K,
    RMSE = acc[2, "RMSE"],
    MAE = acc[2, "MAE"]
  )
})

do.call(rbind, results)
```
Now K=2 is performing the best, try auto arima since arima parameters may change with K.
```{r}
# Generate training regressors
fourier_train <- fourier(ts_RRS_arima, K = 2)

# Fit model
fit_auto <- auto.arima(ts_RRS_arima,
                       xreg = fourier_train,
                       seasonal = FALSE)

# Generate future regressors
fourier_test <- fourier(ts_RRS_arima, K = 2, h = n_test)

# Forecast WITH xreg
fc_auto <- forecast(fit_auto,
                    xreg = fourier_test,
                    h = n_test)

# Accuracy
accuracy(fc_auto, test_data$RRS)
```
**Update the fourier model with best K and ARIMA parameters**
```{r}
#ARIMA + Fourier forecast and accuracy
fourier_train <- fourier(ts_RRS_arima, K = 2)
ARIMA_Fourier_RRS <- Arima(ts_RRS_arima,
                            order = c(1, 1, 1),
                            include.drift = FALSE,
                            xreg = fourier_train)

fourier_test <- fourier(ts_RRS_arima, K = 2, h = n_test)
forecast_fourier <- forecast(ARIMA_Fourier_RRS, xreg = fourier_test, h = n_test)
accuracy(forecast_fourier, test_data$RRS)
```
```

