# NNETAR
Neural network model (NNETAR) to capture non-linear trends and forecast price picks.

```{r}
library(forecast)
RRS_train <- read.csv("../Data/train_period_clean/data_train.csv")
RRS_test <- read.csv("../Data/test_period_clean/data_test.csv")
ts_RRS_train <- ts(RRS_train$RRS,
                   frequency = 24)
ts_RRS_test <- ts(RRS_test$RRS,
                   frequency = 24)
```

```{r}
# fit model
RRS_nnetar_01 <- nnetar(ts_RRS_train)

# forecast
RRS_nnetar_fc <- forecast(RRS_nnetar_01, h = 2352)

plot(RRS_nnetar_fc)
```

```{r}
# Accuracy
accuracy(RRS_nnetar_fc, RRS_test$RRS)
```

The auto model generates m = 24, p (nonseasonal lag) = 28, and P (seasonal lag) = 1, size = 14. This size (number of layers in this network) is unnecessarily high for a simple model, which can be furthur tuned if chosen.

# TABATS
For TABATS, we are using multiseasonal time series.
```{r}
mtts_RRS_train <- msts(
  RRS_train$RRS,
  seasonal.periods = c(24, 168)
)
mtts_RRS_test <- msts(
  RRS_test$RRS,
  seasonal.periods = c(24, 168)
)

# Fit TBATS model
RRS_tbats_01 <- tbats(mtts_RRS_train, lambda = "auto")

# Forecast
RRS_tbats_fc <- forecast(RRS_tbats_01, h = 2352)

# Plot forecast
plot(RRS_tbats_fc,include = 0)
lines(mtts_RRS_test, col = "red")
```
```{r}
# Accuracy
accuracy(RRS_tbats_fc, RRS_test$RRS)
```

# Seasonal Naive as a baseline comparison
Daily seasonal naive
```{r}
# Seasonal naive (daily pattern)
RRS_snaive_fc <- snaive(ts_RRS_train, h = 2352)

# Plot
plot(RRS_snaive_fc)
lines(ts_RRS_test, col = "red")

# Accuracy
accuracy(RRS_snaive_fc,  RRS_test$RRS)
```



